---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(deglr)
library(glmnet)
library(survival)
library(tidyverse)
```

## Setup

Here we set the number of observations from the target and external populations
as well as the number of predictors:
```{r}
n_tar = 50;
n_ext = 1000;
d = 5;
```

We will generate the predictors from multivariate normal distributions with mean
zero and covariance matrices sampled from a Wishart distribution. You can increase
the `wishart_precision` to force the diagonals closer to 1, and you can change
`correlation` to increase or decrease the correlation between predictors. You 
can change `scale_target` to modify the relative values of covariance matrices
between the populations; a value of 1 means they share the same covariance matrix;
a value less than 1 means that the target population has less variability. 

```{r}
wishart_precision = 10; 
correlation = 0.15
scale_target = 0.45;
set.seed(1);
cov_tar = rWishart(1, df = d * wishart_precision, (correlation + diag(1 - correlation, d)) / (d * wishart_precision))[, , 1]
cov_ext <- cov_tar * scale_target
```

Now generate the predictors:

```{r}
xtar <- matrix(rnorm(n_tar * d), ncol = d) %*%  chol(cov_tar)
xext <- matrix(rnorm(n_ext * d), ncol = d) %*%  chol(cov_ext)
```

Now calculate the linear predictors ($X_i^\top \beta$ or $X_i^\top(\beta+\gamma)$): 
```{r}
betas <- numeric(d);
gammas <- 0.05*seq(-10, 10, length = d)
lp_tar <- drop(xtar %*% betas);
lp_ext <- drop(xext %*% (betas + gammas))
```

We will use these linear predictors for each of the families below. 

## `binomial` family

Generate binomial outcomes:

```{r}
set.seed(1);
ytar <- 1 * (lp_tar + rlogis(n_tar, scale = pi / sqrt(3)) > 0)
yext <- 1 * (lp_ext + rlogis(n_ext, scale = pi / sqrt(3)) > 0)
```

Both of these should give equivalent results to our method using $\lambda = 0$:
```{r}
glm(ytar ~ xtar, family = "binomial") %>% coef() 
glmnet(x = xtar, y = ytar, alpha = 0,  family = "binomial", lambda = 0) %>% coef() %>% drop()
```

This is the DEGLR method with $\lambda \approx 0$, $\lambda$ very large:
```{r}
deglr(xtar = xtar, ytar = ytar, xext = xext, yext = yext, family = "binomial", 
      standardize = TRUE, 
      nlambda = 50,
      lambda.min.ratio = 1e-20,
      return_only_at = c(1e-16, 1e3),
      thresh = 1e-7, maxit = 1e5)$beta_hat

```

This should give equivalent results to $\lambda$ very large:
```{r}
glm(c(ytar, yext) ~ rbind(xtar, xext), family = "binomial") %>% coef()
```


## `cox` family

We generate time-to-event outcomes with some censoring:

```{r}
set.seed(1);
time_ytar <- rexp(n_tar, exp(lp_tar))
time_yext <- rexp(n_ext, exp(lp_ext))
cens_ytar <- rexp(n_tar, 0.25 * mean(exp(lp_tar)))
cens_yext <- rexp(n_ext, 0.25 * mean(exp(lp_tar)))
ytar = Surv(pmin(time_ytar, cens_ytar), 1*(time_ytar < cens_ytar))
yext = Surv(pmin(time_yext, cens_yext), 1*(time_yext < cens_yext))
```

Both of these should give equivalent results to our method using $\lambda = 0$:
```{r}
coxph(ytar ~ xtar) %>% coef()
glmnet(x = xtar, y = ytar, alpha = 0,  family = "cox", lambda = 0) %>% coef() %>% drop()
```


This is the DEGLR method with $\lambda \approx 0$, $\lambda$ very large:
```{r}
deglr(xtar = xtar, ytar = ytar, xext = xext, yext = yext, family = "cox", 
      standardize = TRUE, 
      nlambda = 50,
      lambda.min.ratio = 1e-20,
      return_only_at = c(1e-16, 1e3),
      thresh = 1e-7, maxit = 1e5)$beta_hat

```

This should give equivalent results to $\lambda$ very large:
```{r}
coxph(c(ytar, yext) ~ rbind(xtar, xext)) %>% coef()
```

```{r}
if(0) {
  
  
  nfolds = 5;
  ncvreps = 10;
  nlambda = 20;
  lambda.min.ratio = 1e-8;
  fixed_lambda_seq = NULL
  L = NULL;
  standardize = TRUE;
  thresh = 1e-7;
  maxit = 1e5
  
  foo = cv_deglr(xtar = xtar, 
                 ytar = ytar, 
                 yext = yext, 
                 xext = xext, 
                 family = family, 
                 nfolds = nfolds, 
                 ncvreps = ncvreps, 
                 nlambda = nlambda)
  
  foo$beta_hat
  sqrt(mean(foo$beta_hat^2))
  sqrt(mean(coef(glmnet(x = xtar, y = ytar, alpha = 0,  family = family, lambda = 0))^2))
  foo$lambda_opt
}

if(0) {
  foo = deglr(xtar = xtar, ytar = ytar, xext = xext, yext = yext, family = family, lambda = c(0), standardize = TRUE, thresh = 1e-7, maxit = 1e5)
  lp_aug = c(numeric(n_tar),(cbind(1, xext) %*% foo$gamma_hat))
  if(family == "binomial") {
    glm(c(ytar, yext) ~ rbind(xtar, xext) + offset(lp_aug), family = "binomial") %>% coef() %>% print()
  } else if (family == "cox") {
    coxph(c(ytar, yext) ~ rbind(xtar, xext) + offset(lp_aug)) %>% coef() %>% print()
  }
  foo$beta_hat
}
```




